---
title: "0404"
author: "Kexin Guo"
date: "2025-02-20"
output: pdf_document
---

```{r}
library(tidyverse)
```

```{r}
# load in metadata and filter only for low and high pro
# previous before 09102025
# metadata <- read_csv("data/ibd_sparc.csv")
# updated on 0910
metadata = read_csv("data/meta_data0829.csv")

metadata <- metadata %>%
  mutate(
    patient_id = DEIDENTIFIED_MASTER_PATIENT_ID,
    sample_id  = sample_id_y                     # or whatever column you're using
   )
#head(metadata)
temp <- metadata %>%
  dplyr::select(patient_id, sample_id, pro_grp)
head(temp)
# read in microbiome data
species_count <- read_csv("data/species_specimen.nreads.csv")
#head(species_count)
#species_count$specimen

n <- species_count$tax_name
species_count2 <- as.data.frame(t(species_count[,-1]))
colnames(species_count2) <- n
#head(species_count2)

# # now filter for only high and low pro patients
# species_count3 <- species_count2 %>%
#   rownames_to_column("sample_id") %>%
#   merge(temp, by="sample_id") %>%
#   #dplyr::select(-pro_grp, -shannon, -invsimp, -richness) %>%
#   arrange(sample_id) %>%
#   dplyr::filter(pro_grp == "Low-PRO" | pro_grp == "High-PRO") %>%
#   dplyr::select(-pro_grp, -patient_id) %>%
#   column_to_rownames("sample_id")
# head(species_count3)
```


```{r}
# Need to impute zeros
library(zCompositions)

zcomp <- cmultRepl(species_count2, method = "BL", z.delete=F)

# next let's perform CLR transformation to for scale invariant 
log_zcomp <- log(zcomp)
clr_zcomp <- apply(log_zcomp, 2, function(x) x - rowMeans(log_zcomp))

clr_zcomp <- as.data.frame((clr_zcomp))
```
```{r}
#head(clr_zcomp, 10)
# now filter for only high and low pro patients
clr_zcomp1 <- clr_zcomp %>%
  rownames_to_column("sample_id") %>%
  merge(temp, by="sample_id") %>%
  #dplyr::select(-pro_grp, -shannon, -invsimp, -richness) %>%
  arrange(sample_id) %>%
  distinct(sample_id, .keep_all = TRUE) %>%
  dplyr::filter(pro_grp == "Low-PRO" | pro_grp == "High-PRO") %>%
  dplyr::select(-pro_grp, -patient_id) %>%
  column_to_rownames("sample_id")
#head(clr_zcomp1)
```


vs 0219 veision(enriched) Get those species from ancom_fdr dataset that were differentially abundant with lfc > 2
```{r}
ancom_fdr <- read_csv("data/ancom_fdr.csv")

# # take only species with lfc>2
# ancom_fdr_filt <- ancom_fdr %>%
#   dplyr::filter(lfc < -2)

# now select species in clr_zcomp that are DA by ancom-bc
species <- as.vector(ancom_fdr$species)
species
```
select only DA species from dataset
```{r}
enriched_microbes <- clr_zcomp1 %>%
  rownames_to_column("sample_id") %>%
  arrange(sample_id) %>%
  column_to_rownames("sample_id")

# take DA microbes from ANCOM-BC, start with only enriched microbes first (lfc > 2)
# f <- ancom_fdr %>%
#   filter(lfc < -2)
# 
# f <- f$species

y <- enriched_microbes %>%
  #dplyr::select(matches(ancom_fdr$species)) %>%
  #dplyr::filter(rownames(enriched_microbes) %in% f) %>%
  as.matrix()
```

```{r}
head(enriched_microbes)
dim(enriched_microbes)
dim(y)
```

load gene dataset
```{r}
# use gene data as the x matrix
temp2 <- read_csv("data/filtered_count_matrix0910.csv") # Kexin, you will read in the gene dataset here
#head(temp2)

temp3 <- readxl::read_xlsx("data/ibd_sparc_rnaseq.xlsx")
#head(temp3)
```
```{r}
temp3 <- temp3 %>%
  dplyr::select(DEIDENTIFIED_MASTER_PATIENT_ID, CHARACTERISTICS_BIO_MATERIAL, REQUESTED.FILE.NAME) %>%
  distinct(DEIDENTIFIED_MASTER_PATIENT_ID, .keep_all = TRUE) %>%
  mutate(REQUESTED.FILE.NAME = str_extract(REQUESTED.FILE.NAME, "^[^_]+")) %>%
  mutate(CHARACTERISTICS_BIO_MATERIAL = case_when(
    CHARACTERISTICS_BIO_MATERIAL %in% c("At 20 cm", "Rectum", "Sigmoid Colon", "Descending Colon") ~ "left_colon",
    CHARACTERISTICS_BIO_MATERIAL %in% c("Ileum", "Cecum", "Transverse Colon", "Ascending Colon", "Other") ~ "ileum",
    TRUE ~ CHARACTERISTICS_BIO_MATERIAL
  ))
  # mutate(CHARACTERISTICS_BIO_MATERIAL = recode(CHARACTERISTICS_BIO_MATERIAL,
  #                                              "At 20 cm" = "left_colon",
  #                                              "Rectum" = "left_colon",
  #                                              "Sigmoid Colon" = "left_colon",
  #                                              "Descending Colon" = "left_colon",
  #                                              "Ileum" = "ileum",
  #                                              "Cecum" = "ileum",
  #                                              "Transverse Colon" = "ileum",
  #                                              "Ascending Colon" = "ileum",
  #                                              "Other" = "ileum"))
# sample is the sample_id for metabolomics data (or RNAseq data in our case), patient_id will be the same across the microbiome, metabolomics, and RNAseq data
#colnames(temp3)
temp3 = temp3 %>%
  rename_with(~ c("patient_id", "site", "sample_id"),
              .cols = c("DEIDENTIFIED_MASTER_PATIENT_ID", "CHARACTERISTICS_BIO_MATERIAL", "REQUESTED.FILE.NAME"))

temp3$site = factor(temp3$site, levels = c("ileum", "left_colon"))
head(temp3)
dim(temp3)
```

```{r}
temp4 <- temp2 %>%
  dplyr::select(-1) %>%
  #rename(sample_id = sample_id_x) %>%
  rename_with(~ "sample_id", .cols = contains("sample_id")) %>%
  inner_join(temp3, by = "sample_id") 
temp4_ileum <- temp4 %>%
  dplyr::filter(site == "ileum") %>%
  dplyr::select(sample_id, patient_id, everything()) 
temp4_left_colon <- temp4 %>%
  dplyr::filter(site == "left_colon") %>%
  dplyr::select(sample_id, patient_id, everything()) 
# head(temp4_ileum)
# head(temp4_left_colon)
```
```{r}
temp <- temp %>%
  mutate(patient_id = as.character(patient_id))

temp5_ileum <- temp4_ileum %>%
  dplyr::rename(sample_gene = sample_id) %>%
  mutate(patient_id = as.character(patient_id)) %>%
  inner_join(temp, by="patient_id") %>%
  distinct() %>%
  arrange(sample_id) 
#head(temp5_ileum, 3)
dim(temp5_ileum)
temp5_ileum_low = temp5_ileum %>%
  dplyr::filter(pro_grp == "Low-PRO") %>%
  dplyr::select(-pro_grp, -patient_id, -sample_gene) %>%
  relocate(sample_id, .before = everything())
dim(temp5_ileum_low)
#head(temp5_ileum_low, 3)
temp5_ileum_high = temp5_ileum %>%
  dplyr::filter(pro_grp == "High-PRO") %>%
  dplyr::select(-pro_grp, -patient_id, -sample_gene) %>%
  relocate(sample_id, .before = everything())
dim(temp5_ileum_high)
```
```{r}
## in total 60+ to 40+
temp5_left_colon <- temp4_left_colon %>%
  dplyr::rename(sample_gene = sample_id) %>%
  mutate(patient_id = as.character(patient_id)) %>%
  inner_join(temp, by="patient_id") %>%
  distinct() %>%
  arrange(sample_id) 

temp5_left_colon_low = temp5_left_colon %>%
  dplyr::filter(pro_grp == "Low-PRO") %>%
  dplyr::select(-pro_grp, -patient_id, -sample_gene) %>%
  relocate(sample_id, .before = everything())
temp5_left_colon_high = temp5_left_colon %>%
  dplyr::filter(pro_grp == "High-PRO") %>%
  dplyr::select(-pro_grp, -patient_id, -sample_gene) %>%
  relocate(sample_id, .before = everything())
#head(temp5_left_colon, 3)
dim(temp5_left_colon)
dim(temp5_left_colon_low)
dim(temp5_left_colon_high)
```


eliminate variables(genes) that have low variance and are unlikely to be associated with the outcome
```{r}
library(caret)

# xx <- temp5 %>%
#   column_to_rownames("gene") %>%
#   t %>%
#   as.data.frame() %>%
#   rownames_to_column("sample_id") %>%
#   arrange(sample_id) %>%
#   column_to_rownames("sample_id") 
xx_ileum_low <- temp5_ileum_low 
#%>%
#   column_to_rownames("sample_id") %>% # Set sample_id as rownames
#   t() %>% # Transpose the data
#   as.data.frame()

nzv1_low <- caret::nearZeroVar(xx_ileum_low, saveMetrics = T)
#nzv[nzv[,"zeroVar"] > 0, ]
nzv1_low <- nzv1_low[nzv1_low[,"zeroVar"] + nzv1_low[,"nzv"] > 0, ]

nzv1_low <- nzv1_low %>%
  rownames_to_column("col")

drop1_low <- nzv1_low$col

# remove nzv columns
x2_ileum_low <- xx_ileum_low[, !colnames(xx_ileum_low) %in% drop1_low] %>% # Drop columns with low variance
  #t() %>% # Transpose back to original format
  as.data.frame()
dim(x2_ileum_low)
```
```{r}
xx_ileum_high <- temp5_ileum_high 
nzv1_high <- caret::nearZeroVar(xx_ileum_high, saveMetrics = T)
nzv1_high <- nzv1_high[nzv1_high[,"zeroVar"] + nzv1_high[,"nzv"] > 0, ]

nzv1_high <- nzv1_high %>%
  rownames_to_column("col")

drop1_high <- nzv1_high$col

# remove nzv columns
x2_ileum_high <- xx_ileum_high[, !colnames(xx_ileum_high) %in% drop1_high] %>% # Drop columns with low variance
  as.data.frame()
dim(x2_ileum_high)
```

```{r}
xx_left_colon_low <- temp5_left_colon_low 
nzv2_low <- caret::nearZeroVar(xx_left_colon_low, saveMetrics = T)
nzv2_low <- nzv2_low[nzv2_low[,"zeroVar"] + nzv2_low[,"nzv"] > 0, ]

nzv2_low <- nzv2_low %>%
  rownames_to_column("col")

drop2_low <- nzv2_low$col

# remove nzv columns
x2_left_colon_low <- xx_left_colon_low[, !colnames(xx_left_colon_low) %in% drop2_low] %>% # Drop columns with low variance
  #t() %>% # Transpose back to original format
  as.data.frame()
dim(x2_left_colon_low) # 54 to 38
```
```{r}
xx_left_colon_high <- temp5_left_colon_high 
nzv2_high <- caret::nearZeroVar(xx_left_colon_high, saveMetrics = T)
nzv2_high <- nzv2_high[nzv2_high[,"zeroVar"] + nzv2_high[,"nzv"] > 0, ]

nzv2_high <- nzv2_high %>%
  rownames_to_column("col")

drop2_high <- nzv2_high$col

# remove nzv columns
x2_left_colon_high <- xx_left_colon_high[, !colnames(xx_left_colon_high) %in% drop2_high] %>% # Drop columns with low variance
  #t() %>% # Transpose back to original format
  as.data.frame()
dim(x2_left_colon_high) # 8 to 6
```

#combine high and low for both sites
```{r}
# Step 1: Find common genes
common_genes <- intersect(colnames(x2_ileum_high), colnames(x2_ileum_low))
length(common_genes)
# Step 2: Subset the dataframes to common genes
x2_ileum_high_common <- x2_ileum_high[, common_genes]
x2_ileum_low_common  <- x2_ileum_low[, common_genes]

# Step 3: Combine the two dataframes (row bind)
x2_ileum_combined <- rbind(x2_ileum_high_common, x2_ileum_low_common)

# Step 4: Create phenotype vector
phenotype <- c(rep(1, nrow(x2_ileum_high_common)),   # 1 for all high_PRO samples
               rep(0, nrow(x2_ileum_low_common)))    # 0 for all low_PRO samples

```

```{r}
# Step 1: Find common genes
common_genes_left <- intersect(colnames(x2_left_colon_high), colnames(x2_left_colon_low))
length(common_genes_left)
# Step 2: Subset both dataframes to the common genes
x2_left_colon_high_common <- x2_left_colon_high[, common_genes_left]
x2_left_colon_low_common  <- x2_left_colon_low[, common_genes_left]

# Step 3: Combine the two dataframes (rows = samples)
x2_left_colon_combined <- rbind(x2_left_colon_high_common, x2_left_colon_low_common)

# Step 4: Create binary phenotype vector (1 = high PRO, 0 = low PRO)
phenotype_left_colon <- c(rep(1, nrow(x2_left_colon_high_common)),
                          rep(0, nrow(x2_left_colon_low_common)))
```

## updated on 09102025 fot generating files for 0428sparsecca.ipynb
```{r}
# ======================
# SPLIT BY PRO GROUP
# ======================

# -- ILEUM --
x.ileum.high <- x2_ileum_high
x.ileum.low  <- x2_ileum_low

y.ileum.high <- y %>%
  as.data.frame() %>%
  mutate(sample_id = rownames(.)) %>%
  relocate(sample_id, .before = everything()) %>%
  dplyr::filter(sample_id %in% x.ileum.high$sample_id) %>%
  mutate(sample_id = factor(sample_id, levels = x.ileum.high$sample_id)) %>%
  arrange(sample_id)

y.ileum.low <- y %>%
  as.data.frame() %>%
  mutate(sample_id = rownames(.)) %>%
  relocate(sample_id, .before = everything()) %>%
  dplyr::filter(sample_id %in% x.ileum.low$sample_id) %>%
  mutate(sample_id = factor(sample_id, levels = x.ileum.low$sample_id)) %>%
  arrange(sample_id)

# -- LEFT COLON --
x.left.colon.high <- x2_left_colon_high
x.left.colon.low  <- x2_left_colon_low

y.left.colon.high <- y %>%
  as.data.frame() %>%
  mutate(sample_id = rownames(.)) %>%
  relocate(sample_id, .before = everything()) %>%
  dplyr::filter(sample_id %in% x.left.colon.high$sample_id) %>%
  mutate(sample_id = factor(sample_id, levels = x.left.colon.high$sample_id)) %>%
  arrange(sample_id)

y.left.colon.low <- y %>%
  as.data.frame() %>%
  mutate(sample_id = rownames(.)) %>%
  relocate(sample_id, .before = everything()) %>%
  dplyr::filter(sample_id %in% x.left.colon.low$sample_id) %>%
  mutate(sample_id = factor(sample_id, levels = x.left.colon.low$sample_id)) %>%
  arrange(sample_id)
```
```{r}
stopifnot(all(x.ileum.high$sample_id == y.ileum.high$sample_id))
stopifnot(all(x.left.colon.high$sample_id == y.left.colon.high$sample_id))
```
```{r}
dim(x.ileum.high)
dim(x.ileum.low)
dim(y.ileum.high)
dim(y.ileum.low)
```

```{r}
# Helper function
convert_matrix <- function(df) {
  rownames(df) <- df$sample_id
  df <- df[, -1]
  as.matrix(df)
}

x.ileum.high         <- convert_matrix(x.ileum.high)
x.ileum.low          <- convert_matrix(x.ileum.low)
x.left.colon.high    <- convert_matrix(x.left.colon.high)
x.left.colon.low     <- convert_matrix(x.left.colon.low)

y.ileum.high         <- convert_matrix(y.ileum.high)
y.ileum.low          <- convert_matrix(y.ileum.low)
y.left.colon.high    <- convert_matrix(y.left.colon.high)
y.left.colon.low     <- convert_matrix(y.left.colon.low)
```

```{r}
# Save X
write.csv(x.ileum.high,        "x565.ileum.high.csv",        row.names = TRUE)
write.csv(x.ileum.low,         "x565.ileum.low.csv",         row.names = TRUE)
write.csv(x.left.colon.high,   "x565.left.colon.high.csv",   row.names = TRUE)
write.csv(x.left.colon.low,    "x565.left.colon.low.csv",    row.names = TRUE)

# Save Y
write.csv(y.ileum.high,        "y565.ileum.high.csv",        row.names = TRUE)
write.csv(y.ileum.low,         "y565.ileum.low.csv",         row.names = TRUE)
write.csv(y.left.colon.high,   "y565.left.colon.high.csv",   row.names = TRUE)
write.csv(y.left.colon.low,    "y565.left.colon.low.csv",    row.names = TRUE)

```

## previous code on generating high low combined datasets
```{r}
# y.ileum <- y %>%
#   as.data.frame(y) %>%
#   mutate(sample_id = rownames(y)) %>%
#   relocate(sample_id, .before = everything()) %>%
#   #rownames_to_column(var = "sample_id") %>%
#   dplyr::filter(sample_id %in% x2_ileum_combined$sample_id) %>%
#   arrange(sample_id) 
#   #tibble::column_to_rownames("sample_id")
x.ileum <- x2_ileum_combined 
y.ileum <- y %>%
  as.data.frame() %>%
  mutate(sample_id = rownames(.)) %>%
  relocate(sample_id, .before = everything()) %>%
  dplyr::filter(sample_id %in% x.ileum$sample_id)   # Keep only samples in x
  # Do NOT arrange here yet

# Now, **reorder y** to match **the sample order of x**
y.ileum <- y.ileum %>%
  mutate(sample_id = factor(sample_id, levels = x.ileum$sample_id)) %>%  # keep original x order
  arrange(sample_id) 
# y.ileum.high <- y %>%
#   as.data.frame(y) %>%
#   mutate(sample_id = rownames(y)) %>%
#   relocate(sample_id, .before = everything()) %>%
#   dplyr::filter(sample_id %in% x2_ileum_high$sample_id) %>%
#   arrange(sample_id) 
# x.ileum.high <- x2_ileum_high 

#%>%as.matrix() 
# y.left.colon <- y %>%
#   as.data.frame(y) %>%
#   mutate(sample_id = rownames(y)) %>%
#   relocate(sample_id, .before = everything()) %>%
#   dplyr::filter(sample_id %in% x2_left_colon_combined$sample_id) %>%
#   arrange(sample_id) 
x.left.colon <- x2_left_colon_combined
y.left.colon <- y %>%
  as.data.frame() %>%
  mutate(sample_id = rownames(.)) %>%
  relocate(sample_id, .before = everything()) %>%
  dplyr::filter(sample_id %in% x.left.colon$sample_id) 

y.left.colon <- y.left.colon %>%
  mutate(sample_id = factor(sample_id, levels = x.left.colon$sample_id)) %>%
  arrange(sample_id)

# y.left.colon.high <- y %>%
#   as.data.frame(y) %>%
#   mutate(sample_id = rownames(y)) %>%
#   relocate(sample_id, .before = everything()) %>%
#   dplyr::filter(sample_id %in% x2_left_colon_high$sample_id) %>%
#   arrange(sample_id) 
# x.left.colon.high <- x2_left_colon_high 
#%>%as.matrix() 
```
```{r}
dim(y.ileum)
dim(y.left.colon)
```

```{r}
# make sure that all rows in x match all rows in y
stopifnot(all(x.ileum$sample_id == y.ileum$sample_id))
stopifnot(all(x.left.colon$sample_id == y.left.colon$sample_id))
# stopifnot(all(x.ileum.high$sample_id == y.ileum.high$sample_id))
# stopifnot(all(x.left.colon.high$sample_id == y.left.colon.high$sample_id))
```
```{r}
# head(x2_ileum_high_common)
# head(x2_ileum_combined)
```

### lasso regression microbe-wise
```{r}
rownames(x.ileum) <- x.ileum$sample_id  # Set sample_id as row names
x.ileum <- x.ileum[, -1]  # Remove the sample_id column
# Convert to a numeric matrix
x.ileum <- as.matrix(x.ileum)
rownames(x.left.colon) <- x.left.colon$sample_id  # Set sample_id as row names
x.left.colon <- x.left.colon[, -1]  # Remove the sample_id column
# Convert to a numeric matrix
x.left.colon <- as.matrix(x.left.colon)
dim(x.ileum)
dim(x.left.colon)

# rownames(x.ileum.high) <- x.ileum.high$sample_id  # Set sample_id as row names
# x.ileum.high <- x.ileum.high[, -1]  # Remove the sample_id column
# # Convert to a numeric matrix
# x.ileum.high <- as.matrix(x.ileum.high)
# rownames(x.left.colon.high) <- x.left.colon.high$sample_id  # Set sample_id as row names
# x.left.colon.high <- x.left.colon.high[, -1]  # Remove the sample_id column
# # Convert to a numeric matrix
# x.left.colon.high <- as.matrix(x.left.colon.high)
# dim(x.ileum.high)
# dim(x.left.colon.high)
```
```{r}
rownames(y.ileum) <- y.ileum$sample_id  # Set sample_id as row names
y.ileum <- y.ileum[, -1]  # Remove the sample_id column
# Convert to a numeric matrix
y.ileum <- as.matrix(y.ileum)
rownames(y.left.colon) <- y.left.colon$sample_id  # Set sample_id as row names
y.left.colon <- y.left.colon[, -1]  # Remove the sample_id column
# Convert to a numeric matrix
y.left.colon <- as.matrix(y.left.colon)
dim(y.ileum)
dim(y.left.colon)

# rownames(y.ileum.high) <- y.ileum.high$sample_id  # Set sample_id as row names
# y.ileum.high <- y.ileum.high[, -1]  # Remove the sample_id column
# # Convert to a numeric matrix
# y.ileum.high <- as.matrix(y.ileum.high)
# rownames(y.left.colon.high) <- y.left.colon.high$sample_id  # Set sample_id as row names
# y.left.colon.high <- y.left.colon.high[, -1]  # Remove the sample_id column
# # Convert to a numeric matrix
# y.left.colon.high <- as.matrix(y.left.colon.high)
# dim(y.ileum.high)
# dim(y.left.colon.high)
# #colnames(y.ileum.high)
```
```{r}
# Save X matrices
write.csv(x.ileum, "x565.ileum.csv", row.names = TRUE)
write.csv(x.left.colon,  "x565.left.colon.csv",  row.names = TRUE)

# Save Y matrices
write.csv(y.ileum, "y565.ileum.csv", row.names = TRUE)
write.csv(y.left.colon,  "y565.left.colon.csv",  row.names = TRUE)

write.csv(phenotype, "phenotype.ileum", row.names = TRUE)
write.csv(phenotype_left_colon, "phenotype.left.colon", row.names = TRUE)
```

update for 0428 on using SmCCNet
# SmCCNet
```{r}
x.ileum = read.csv("x565.ileum.csv", row.names = 1)
y.ileum <- read.csv("y565.ileum.csv", row.names = 1)
phenotype = read.csv("phenotype.ileum", row.names = 1)$x

x.left.colon = read.csv("x565.left.colon.csv", row.names = 1)
y.left.colon = read.csv("y565.left.colon.csv", row.names = 1)
phenotype_left_colon = read.csv("phenotype.left.colon", row.names = 1)$x
```

```{r}
library(SmCCNet)
```

```{r}
table(phenotype)
table(phenotype_left_colon)
```
```{r}
sum(is.na(x.ileum))   # Count missing values
sum(is.infinite(as.matrix(x.ileum)))   # Count Inf values

sum(is.na(y.ileum)) 
sum(is.infinite(as.matrix(y.ileum)))
```
```{r}
# Select top 1500 genes by variance
gene_variance <- apply(x.ileum, 2, var)
top_genes <- names(sort(gene_variance, decreasing = TRUE))[1:1500]
x_ileum_topvar <- x.ileum[, top_genes]
x_ileum_topvar <- as.matrix(x_ileum_topvar)
dim(x_ileum_topvar)
length(phenotype)
# Make sure sample orders match
stopifnot(rownames(x_ileum_topvar) == rownames(y.ileum))
stopifnot(length(phenotype) == nrow(x_ileum_topvar))
```
```{r}
head(phenotype)
typeof(phenotype)
length(phenotype)
```
## ileum 
```{r}
# Ensure phenotype is factor for binary case
phenotype <- as.factor(phenotype)
y.ileum = as.matrix(y.ileum)

# Run fastAutoSmCCNet for ileum
result_ileum <- suppressWarnings(fastAutoSmCCNet(
  X = list(x_ileum_topvar, y.ileum),   # two omics layers
  Y = phenotype,          # binary labels
  Kfold = 3,                    # 3-fold cross-validation
  subSampNum = 100,             # number of subsampling (higher = more robust)
  DataType = c('Gene', 'Microbe'),  # give nice names to datasets
  saving_dir = "0428_ileum_results",  # create a new folder for results
  EvalMethod = 'auc',           # use AUC for binary evaluation
  summarization = 'NetSHy',     # NetSHy for module summarization
  CutHeight = 1 - 0.1^10,       # very low cut height, keeping modules tight
  BetweenShrinkage = 5,         # shrinkage between omics
  ncomp_pls = 3                 # number of PLS components
))
```
## left colon
```{r}
# Top 1500 genes
gene_variance_left <- apply(x.left.colon, 2, var)
top_genes_left <- names(sort(gene_variance_left, decreasing = TRUE))[1:1500]
x_left_colon_topvar <- x.left.colon[, top_genes_left]

# Ensure numeric matrices
x_left_colon_topvar <- as.matrix(x_left_colon_topvar)
y.left.colon <- as.matrix(y.left.colon)
```
```{r}
# Run fastAutoSmCCNet
result_left_colon <- suppressWarnings(
  fastAutoSmCCNet(
    X = list(x_left_colon_topvar, y.left.colon),
    Y = as.factor(phenotype_left_colon),
    Kfold = 3,
    subSampNum = 100,
    DataType = c('Gene', 'Microbe'),
    saving_dir = "0428_leftcolon_results",
    EvalMethod = 'auc',
    summarization = 'NetSHy',
    CutHeight = 1 - 0.1^10,
    BetweenShrinkage = 5,
    ncomp_pls = 3
  )
)
```

## performance
```{r}
#rm(list = ls())
```


```{r}
load("0428_ileum_results/globalNetwork.Rdata")
str(globalNetwork)
```

```{r}
# library(igraph)
# 
# # Folder where your Rdata files are
# subnetwork_files <- list.files(path = "0428_ileum_results", pattern = "size_.*\\.Rdata", full.names = TRUE)
# 
# # Initialize a list to store subnetworks
# subnetworks <- list()
# 
# # Load all subnetworks
# for (file in subnetwork_files) {
#   before <- ls()
#   load(file)
#   after <- ls()
#   new_object <- setdiff(after, before)
#   net_object <- get(new_object)
#   str(net_object)
#   # Save to list with clean name
#   subnetworks[[file]] <- net_object
#   
#   # Optional: Remove temporary objects
#   rm(list = new_object)
# }
# 
# # Now plot all subnetworks one by one
# par(mfrow = c(2, 3))  # Adjust grid size depending how many modules you have
# for (i in seq_along(subnetworks)) {
#   g <- graph_from_adjacency_matrix(subnetworks[[i]], mode = "undirected", weighted = TRUE)
#   plot(g, main = paste0("Subnetwork ", i), vertex.label.cex = 0.6, vertex.size = 5, edge.width = E(g)$weight * 5)
# }
```

```{r}
names(result_left_colon$Data[[1]])
names(result_left_colon$Data[[2]])
# See top genes/microbes by absolute weight
# top_genes <- sort(abs(gene_weights_left_colon), decreasing = TRUE)
# top_microbes <- sort(abs(microbe_weights_left_colon), decreasing = TRUE)
# 
# # Show top 10
# head(top_genes, 10)
# head(top_microbes, 10)
```

```{r}
# library(ggplot2)
# 
# # Suppose cv_results has columns: GenePenalty, MicrobePenalty, TestAUC
# ggplot(cv_results, aes(x = l1, y = l2, fill = TestMetric)) +
#   geom_tile() +
#   scale_fill_viridis_c(option = "C") +
#   labs(title = "Test AUC across penalty combinations",
#        x = "Gene Penalty",
#        y = "Microbe Penalty") +
#   theme_minimal()
```
```{r}
# See what is inside
names(result_ileum)
result_left_colon$ClusteringModules
```

