---
title: "1108filterDE"
author: "Kexin Guo"
date: "2024-11-05"
output:
  pdf_document: default
  html_document: default
---

# DE analysis including only gene and pro_grp into the model
## prep
```{r}
rm(list = ls())
library(DESeq2)
library(dplyr)
library(data.table)
library(BiocParallel)
register(MulticoreParam(4))
```

## data loading
```{r}
merge_data = fread('merge.csv')
merge_data = merge_data[pro_grp != "", -1, with = FALSE]
merge_data = as.data.frame(merge_data)
#typeof(genes)
head(merge_data)
```

## DESeq2 workflow
```{r}
#count_matrix_gene = fread('count_matrix_gene.csv')
count_matrix_gene = merge_data[, 1:(ncol(merge_data)-109-196355+1)]
head(count_matrix_gene)
count_matrix_gene = data.frame(count_matrix_gene)
#head(count_matrix_gene)
```
### filtering non-coding genes
```{r}
geneid = data.frame(ensembl_gene_id = colnames(count_matrix_gene)[-1])
dim(geneid)
#head(geneid,2)
```

```{r}
## with more unmatched ensembl
## maybe try v86 later
library(EnsDb.Hsapiens.v79)
anno <- ensembldb::select(
  EnsDb.Hsapiens.v79,
  keys = geneid$ensembl_gene_id,
  keytype = "GENEID",
  columns = c("SYMBOL", "GENEBIOTYPE")
)
# length(anno$SYMBOL)
# # Filter out symbols that start with 'RNA5S'
# filtered_anno <- anno[!grepl("^RNA5S", anno$SYMBOL), ]
# 
# # Display the filtered results
# length(filtered_anno$SYMBOL)
# head(anno)

# library(EnsDb.Hsapiens.v86)
# anno <- ensembldb::select(
#   EnsDb.Hsapiens.v86,
#   keys = geneid$ensembl_gene_id,
#   keytype = "GENEID",
#   columns = c("SYMBOL", "GENEBIOTYPE")
# )
head(anno, 3)
dim(anno)
```

```{r}
# library(biomaRt)
# mart <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
# annotations <- getBM(
#   attributes = c("ensembl_gene_id", "hgnc_symbol", "gene_biotype"),
#   filters = "ensembl_gene_id",
#   values = geneid$ensembl_gene_id,
#   mart = mart
#   #mart = useDataset("hsapiens_gene_ensembl", useMart("ENSEMBL_MART_ENSEMBL"))
# )
# head(annotations, 3)
# dim(annotations)
#table(annotations$hgnc_symbol)
```
use anno <- ensembldb::select(
  EnsDb.Hsapiens.v86,
  keys = geneid$ensembl_gene_id,
  keytype = "GENEID",
  columns = c("SYMBOL", "GENEBIOTYPE")
) to get anno

```{r}
table(anno$GENEBIOTYPE)
```

```{r}
# Extract protein-coding genes from the annotation data
# protein_coding_genes <- anno$GENEID[anno$GENEBIOTYPE == "protein_coding"]
protein_coding_genes = anno$GENEID[anno$GENEBIOTYPE %in% c("protein_coding", "lincRNA")]

# Subset `count_matrix_gene` to retain only columns (gene IDs) that are in `protein_coding_genes`
filtered_count_matrix <- count_matrix_gene[, c(TRUE, colnames(count_matrix_gene)[-1] %in% protein_coding_genes)]

# Check the dimensions of the filtered matrix
dim(filtered_count_matrix)
head(filtered_count_matrix)
```
```{r}
write.csv(filtered_count_matrix, "filtered_count_matrix0910.csv")
```

## updated on 8/29/2025 on metadata
```{r}
#meta_data = fread('meta_data.csv')
meta_data = data.frame(sample_id = merge_data$REQUESTED.FILE.NAME, merge_data[, (ncol(merge_data)-108):ncol(merge_data)])

## updated on 8/29(9 patients with >4 SES)
meta_data = meta_data %>%
  dplyr::filter(total_ses_cd_score <= 4)

head(meta_data)
dim(meta_data)
```
```{r}
write.csv(meta_data, "meta_data0829.csv")
```

```{r}
# # count matrix with only gene (no transcript)
# count_matrix_gene = merge_data[, 1:(ncol(merge_data)-109-196355+1)]
# tail(rownames(t(count_matrix_gene)), 10)
# # all meta data
# meta_data = data.frame(sample_id = merge_data$REQUESTED.FILE.NAME, merge_data[, (ncol(merge_data)-108):ncol(merge_data)])
# head(meta_data)
```

### donot need to run everytime (those files are non-filtered)
```{r}
# write.csv(count_matrix_gene, "count_matrix_gene.csv")
# write.csv(meta_data, "meta_data.csv")
```

# updating on 02/08/2026 (use from metadata0829 & filtered_count_matrix0910)
# RUN FROM HERE
```{r}
rm(list = ls())
library(readr)
library(DESeq2)
library(dplyr)
library(data.table)
library(BiocParallel)
library(EnsDb.Hsapiens.v79)
register(MulticoreParam(4))
```
```{r}
meta_data = data.frame(read_csv("meta_data0829.csv"))
filtered_count_matrix = data.frame(read_csv("filtered_count_matrix0910.csv"))
```
```{r}
meta_data = meta_data[, -1]
dim(meta_data)
filtered_count_matrix = filtered_count_matrix[, -1]
dim(filtered_count_matrix)
```

### 08252026 added
```{r}
table(meta_data$CHARACTERISTICS_BIO_MATERIAL,
      meta_data$pro_grp)

length(unique(meta_data$patient_id))
```



## process to separate data by sites
```{r}
meta_data$CHARACTERISTICS_BIO_MATERIAL <- factor(meta_data$CHARACTERISTICS_BIO_MATERIAL)
meta_data$CHARACTERISTICS_BIO_MATERIAL <- recode(meta_data$CHARACTERISTICS_BIO_MATERIAL,
                                          "At 20 cm" = "left_colon",
                                          "Rectum" = "left_colon",
                                          "Sigmoid Colon" = "left_colon",
                                          "Descending Colon" = "left_colon",
                                          "Ileum" = "ileum",
                                          "Cecum" = "ileum",
                                          "Transverse Colon" = "ileum",
                                          "Ascending Colon" = "ileum", 
                                          "Other" = "ileum")
table(meta_data$CHARACTERISTICS_BIO_MATERIAL)
meta_data$CHARACTERISTICS_BIO_MATERIAL = factor(meta_data$CHARACTERISTICS_BIO_MATERIAL, levels = c("ileum", "left_colon"))


meta_data$pro_grp = factor(meta_data$pro_grp, levels = c("High-PRO", "Low-PRO", "Active CD"))
```
```{r}
# Subsetting meta_data by CHARACTERISTICS_BIO_MATERIAL
meta_data_ileum <- subset(meta_data, CHARACTERISTICS_BIO_MATERIAL == "ileum")
meta_data_ileum = meta_data_ileum[!is.na(meta_data_ileum$pro_grp), ]
meta_data_left_colon <- subset(meta_data, CHARACTERISTICS_BIO_MATERIAL == "left_colon")
meta_data_left_colon = meta_data_left_colon[!is.na(meta_data_left_colon$pro_grp), ]

meta_data_ileum = meta_data_ileum %>% distinct(sample_id, .keep_all = TRUE) 
meta_data_left_colon = meta_data_left_colon %>% distinct(sample_id, .keep_all = TRUE) 
dim(meta_data_ileum)
dim(meta_data_left_colon)
```
```{r}
length(meta_data_ileum$sample_id)
length(filtered_count_matrix$sample_id_x)

```

```{r}
#head(count_matrix_gene)
# count_matrix_gene_ileum <- filtered_count_matrix[filtered_count_matrix$sample_id_x %in% meta_data_ileum$sample_id, ]
# count_matrix_gene_left_colon <- filtered_count_matrix[filtered_count_matrix$sample_id_x %in% meta_data_left_colon$sample_id, ]

## 没用的
# count_matrix_gene_ileum = count_matrix_gene_ileum[!duplicated(count_matrix_gene_ileum$sample_id_x)]
# count_matrix_gene_left_colon = count_matrix_gene_left_colon[meta_data_left_colon$sample_id, , drop = FALSE]

count_matrix_gene_ileum = filtered_count_matrix %>%
  dplyr::filter(sample_id_x %in% meta_data_ileum$sample_id) %>%
  distinct(sample_id_x, .keep_all = TRUE) %>%
  arrange(match(sample_id_x, meta_data_ileum$sample_id))

count_matrix_gene_left_colon = filtered_count_matrix %>%
  dplyr::filter(sample_id_x %in% meta_data_left_colon$sample_id) %>%
  distinct(sample_id_x, .keep_all = TRUE) %>%
  arrange(match(sample_id_x, meta_data_left_colon$sample_id))

dim(count_matrix_gene_ileum)
dim(count_matrix_gene_left_colon)
count_matrix_gene_ileum = count_matrix_gene_ileum[, 2: ncol(count_matrix_gene_ileum)]
count_matrix_gene_left_colon = count_matrix_gene_left_colon[, 2: ncol(count_matrix_gene_left_colon)]
head(count_matrix_gene_ileum)
```



```{r}
table(meta_data_left_colon$pro_grp)
length(meta_data_left_colon$pro_grp)
```
```{r}
dim(meta_data_ileum)
dim(count_matrix_gene_ileum)
dim(meta_data_left_colon)
dim(count_matrix_gene_left_colon)
```


## create DESeq object
```{r}
count_matrix_gene_ileum = as.matrix(count_matrix_gene_ileum)
dds_ileum <- DESeqDataSetFromMatrix(countData = t(count_matrix_gene_ileum),
                              colData = meta_data_ileum,
                              design = ~ pro_grp) 
```
```{r}
count_matrix_gene_left_colon = as.matrix(count_matrix_gene_left_colon)
dds_left_colon = DESeqDataSetFromMatrix(countData = t(count_matrix_gene_left_colon), colData = meta_data_left_colon, design = ~ pro_grp)
```

## exploratory
### pre-filtering
```{r}
# ileum
smallestGroupSize = 3
keep1 = rowSums(counts(dds_ileum) >= 10) >= smallestGroupSize  #keep only rows that have a count of at least 10 for a minimal number of samples
dds_ileum = dds_ileum[keep1,]
# left colon
smallestGroupSize = 3
keep2 = rowSums(counts(dds_left_colon) >= 10) >= smallestGroupSize  #keep only rows that have a count of at least 10 for a minimal number of samples
dds_left_colon = dds_left_colon[keep2,]
```

### factor
```{r}
#dds$MACROSCOPIC_APPEARANCE = factor(dds$MACROSCOPIC_APPEARANCE)
dds_ileum$pro_grp = factor(dds_ileum$pro_grp)
dds_left_colon$pro_grp = factor(dds_left_colon$pro_grp)
#dds$anxiety = factor(dds$anxiety)
```

### transform the counts using variance stabilizing transformation (VST)
```{r}
# vsd <- vst(dds, blind = FALSE)
# length(assay(vsd))
# head(assay(vsd)[,1:2], 2)
# dim(assay(vsd))
# #colData(vsd)
```

### Scatterplot of transformed counts ???

### sample distances
```{r}
# sampleDists <- dist(t(assay(vsd)))
```
```{r}
library("pheatmap")
library("RColorBrewer")
```

## differential expression pipeline
```{r}
## Run analysis
dds_ileum <- DESeq(dds_ileum)
dds_left_colon = DESeq(dds_left_colon)
```

### High vs Low contrast
```{r}
library(apeglm)
library(ashr)
resultsNames(dds_ileum)
resultsNames(dds_left_colon)
## remove , lfcThreshold=1.5
# High_pro vs low_pro
ileum_high_vs_low <- results(dds_ileum, contrast=c("pro_grp", "High-PRO", "Low-PRO"))
head(ileum_high_vs_low)
left_colon_high_vs_low = results(dds_left_colon, contrast = c("pro_grp", "High-PRO", "Low-PRO"))
#head(left_colon_high_vs_low)
```
```{r}
# print("number of sig(0.05):")
# resSig_high_low <- subset(res_high_vs_low, padj < 0.1)
# print(sum(res_high_vs_low$padj < 0.1))
print("Gene ordered by log2fold change:")
head(ileum_high_vs_low[order(ileum_high_vs_low$log2FoldChange, decreasing = TRUE), ])
head(left_colon_high_vs_low[order(left_colon_high_vs_low$log2FoldChange, decreasing = TRUE), ])
```

### High vs Active (omitted from year 2026)
```{r}
ileum_high_vs_active <- results(dds_ileum, contrast=c("pro_grp", "High-PRO", "Active CD"))
#head(ileum_high_vs_active)
left_colon_high_vs_active = results(dds_left_colon, contrast = c("pro_grp", "High-PRO", "Active CD"))
```
```{r}
# print("number of sig(0.05):")
# resSig_high_active <- subset(res_high_vs_active, padj < 0.05)
# table(resSig_high_active)
print("Gene ordered by log2fold change:")
head(ileum_high_vs_active[order(ileum_high_vs_active$log2FoldChange, decreasing = TRUE), ])
head(left_colon_high_vs_active[order(left_colon_high_vs_active$log2FoldChange, decreasing = TRUE), ])
```

## annotating and exporting
```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
# BiocManager::install("org.Hs.eg.db")
```
```{r}
columns(org.Hs.eg.db)
```
```{r}
# high vs low result
ens.str <- substr(rownames(ileum_high_vs_low), 1, 15)
ileum_high_vs_low$symbol <- mapIds(org.Hs.eg.db,
                     keys=ens.str,
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
ileum_high_vs_low$entrez <- mapIds(org.Hs.eg.db,
                     keys=ens.str,
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")
print("------------------------------------------")
ens.str <- substr(rownames(left_colon_high_vs_low), 1, 15)
left_colon_high_vs_low$symbol <- mapIds(org.Hs.eg.db,
                     keys=ens.str,
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
left_colon_high_vs_low$entrez <- mapIds(org.Hs.eg.db,
                     keys=ens.str,
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")
```
## latest output files for pathway analysis updated on 02/08/2026 
## (remove the filtering on abs(log2fc) & padj for better performance using gseKEGG()), and should be redone this after pathway analysis
```{r}
## loose the results 
res_low_ileum <- as.data.frame(ileum_high_vs_low[order(-abs(ileum_high_vs_low$log2FoldChange), ileum_high_vs_low$padj),])
library(readr)
library(dplyr)
res_low_ileum <- res_low_ileum %>%
  dplyr::filter(!is.na(symbol) & !is.na(entrez)) %>%
  # dplyr::filter(abs(log2FoldChange) > 2) %>%
  mutate(padj = p.adjust(pvalue, method = "BH")) %>%
  # dplyr::filter(padj < 0.05) %>%
  arrange(padj) 
dim(res_low_ileum)
res_low_ileum
print("------------------------------------------")
res_low_left_colon <- as.data.frame(left_colon_high_vs_low[order(-abs(left_colon_high_vs_low$log2FoldChange), left_colon_high_vs_low$padj),])
#resOrdered_low_DF <- as.data.frame(resOrdered_low)
library(readr)
library(dplyr)
res_low_left_colon <- res_low_left_colon %>%
  dplyr::filter(!is.na(symbol) & !is.na(entrez)) %>%
  # dplyr::filter(abs(log2FoldChange) > 2) %>%
  mutate(padj = p.adjust(pvalue, method = "BH")) %>%
  # dplyr::filter(padj < 0.05) %>%
  arrange(padj) 
dim(res_low_left_colon)
res_low_left_colon
```
## (latest output files for pathway analysis on 09012025)
```{r}
write.csv(res_low_left_colon, file = "1115results_low_high_left_1.0.csv")
write.csv(res_low_ileum, file = "1115results_low_high_ileum_1.0.csv")
```

```{r}
# high vs active result
ens.str <- substr(rownames(ileum_high_vs_active), 1, 15)
ileum_high_vs_active$symbol <- mapIds(org.Hs.eg.db,
                     keys=ens.str,
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
ileum_high_vs_active$entrez <- mapIds(org.Hs.eg.db,
                     keys=ens.str,
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")
print("------------------------------------------")
ens.str <- substr(rownames(left_colon_high_vs_low), 1, 15)
left_colon_high_vs_active$symbol <- mapIds(org.Hs.eg.db,
                     keys=ens.str,
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
left_colon_high_vs_active$entrez <- mapIds(org.Hs.eg.db,
                     keys=ens.str,
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")
```
```{r}
res_active_ileum <- as.data.frame(ileum_high_vs_active[order(-abs(ileum_high_vs_active$log2FoldChange), ileum_high_vs_active$padj),])
res_active_ileum <- res_active_ileum %>%
  filter(!is.na(symbol) & !is.na(entrez)) %>%
  filter(abs(log2FoldChange) > 2) %>%
  mutate(padj = p.adjust(pvalue, method = "BH")) %>%
  filter(padj < 0.05) %>%
  arrange(padj) 
dim(res_active_ileum)
res_active_ileum
print("------------------------------------------")
res_active_left_colon <- as.data.frame(left_colon_high_vs_active[order(-abs(left_colon_high_vs_active$log2FoldChange), left_colon_high_vs_active$padj),])
res_active_left_colon <- res_active_left_colon %>%
  filter(!is.na(symbol) & !is.na(entrez)) %>%
  filter(abs(log2FoldChange) > 2) %>%
  mutate(padj = p.adjust(pvalue, method = "BH")) %>%
  filter(padj < 0.05) %>%
  arrange(padj) 
dim(res_active_left_colon)
res_active_left_colon
```
```{r}
write.csv(res_active_left_colon, file = "1115results_active_high_left_1.0.csv")
write.csv(res_active_ileum, file = "1115results_active_high_ileum_1.0.csv")
```



```{r}
# high vs active result
# ens.str <- substr(rownames(res_high_vs_active), 1, 15)
# res_high_vs_active$symbol <- mapIds(org.Hs.eg.db,
#                      keys=ens.str,
#                      column="SYMBOL",
#                      keytype="ENSEMBL",
#                      multiVals="first")
# res_high_vs_active$entrez <- mapIds(org.Hs.eg.db,
#                      keys=ens.str,
#                      column="ENTREZID",
#                      keytype="ENSEMBL",
#                      multiVals="first")
```
```{r}
# resOrdered_active <- res_high_vs_active[order(abs(res_high_vs_active$log2FoldChange), decreasing = TRUE),]
# #head(resOrdered_active_DF, 5)
# resOrdered_active_DF <- as.data.frame(resOrdered_active)
# resOrdered_active_DF <- resOrdered_active_DF %>%
#   filter(!is.na(symbol) & !is.na(entrez)) %>%
#   filter(abs(log2FoldChange) > 2) %>%
#   mutate(padj = p.adjust(pvalue, method = "BH")) %>%
#   filter(padj < 0.05) %>%
#   arrange(padj) 
# #   mutate(padj = p.adjust(pvalue, method = "BH"))
# head(resOrdered_active_DF, 5)
# dim(resOrdered_active_DF)
```

```{r}
# write.csv(resOrdered_active_DF, file = "results_ordered_high_active.csv")
```


