---
title: "pathway_filter_1.0"
output:
  pdf_document: default
  html_document: default
date: "2024-11-25"
---

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc()
```

## Prep
```{r}
library(tidyverse) # includes ggplot2, for data visualisation. dplyr, for data manipulation.
library(RColorBrewer) # for a colourful plot
library(pheatmap)
library(clusterProfiler) # for PEA analysis
library('org.Hs.eg.db')
library(DOSE)
library(enrichplot) # for visualisations
library(ggupset) # for visualisations
library(dplyr)
```
## Data loading & transformation
### .csv files have been updated since 02/08/2026
### and added "stat" into the df
```{r}
high_low_ileum = read.csv("1115results_low_high_ileum_1.0.csv", header = TRUE, stringsAsFactors = FALSE)
high_low_leftcolon = read.csv("1115results_low_high_left_1.0.csv", header = TRUE, stringsAsFactors = FALSE)

# Rename the columns
colnames(high_low_ileum) <- c("gene", "baseMean", "log2fc", "lfcSE", "stat", "pval", "padj", "gene_symbol", "entrez")
high_low_ileum = high_low_ileum[, c("gene_symbol", "entrez", "stat", "pval", "padj", "log2fc", "lfcSE")]
# Annotate according to differential expression
high_low_ileum <- high_low_ileum %>% mutate(diffexpressed = case_when(
  log2fc > 0 ~ 'UP',
  log2fc < 0 ~ 'DOWN'
))
high_low_ileum


colnames(high_low_leftcolon) <- c("gene", "baseMean", "log2fc", "lfcSE", "stat", "pval", "padj", "gene_symbol", "entrez")
high_low_leftcolon = high_low_leftcolon[, c("gene_symbol", "entrez", "stat", "pval", "padj", "log2fc", "lfcSE")]
high_low_leftcolon <- high_low_leftcolon %>% mutate(diffexpressed = case_when(
  log2fc > 0 ~ 'UP',
  log2fc < 0 ~ 'DOWN'
))
high_low_leftcolon
```
```{r}
high_low_ileum$gene_symbol
print("-------------------------------------------")
high_low_leftcolon$gene_symbol
```
### might skip active comparisons
```{r}
high_active_ileum = read.csv("1115results_active_high_ileum_1.0.csv", header = TRUE, stringsAsFactors = FALSE)
high_active_leftcolon = read.csv("1115results_active_high_left_1.0.csv", header = TRUE, stringsAsFactors = FALSE)

# Rename the columns
colnames(high_active_ileum) <- c("gene", "baseMean", "log2fc", "lfcSE", "stat", "pval", "padj", "gene_symbol", "entrez")
high_active_ileum = high_active_ileum[, c("gene_symbol", "entrez", "pval", "padj", "log2fc", "lfcSE")]
# Annotate according to differential expression
high_active_ileum <- high_active_ileum %>% mutate(diffexpressed = case_when(
  log2fc > 0 ~ 'UP',
  log2fc < 0 ~ 'DOWN'
))
high_active_ileum


colnames(high_active_leftcolon) <- c("gene", "baseMean", "log2fc", "lfcSE", "stat", "pval", "padj", "gene_symbol", "entrez")
high_active_leftcolon = high_active_leftcolon[, c("gene_symbol", "entrez", "pval", "padj", "log2fc", "lfcSE")]
high_active_leftcolon <- high_active_leftcolon %>% mutate(diffexpressed = case_when(
  log2fc > 0 ~ 'UP',
  log2fc < 0 ~ 'DOWN'
))
high_active_leftcolon
```
```{r}
high_active_ileum$gene_symbol
print("-------------------------------------------")
high_active_leftcolon$gene_symbol
```


## run ClusterProfiler
```{r}
length(high_low_ileum$gene_symbol)
length(high_low_leftcolon$gene_symbol)

```

```{r}
enrich_low_ileum <- enrichGO(
  gene = high_low_ileum$gene_symbol,   # DE gene list
  OrgDb = org.Hs.eg.db,          # Human gene annotation database
  keyType = "SYMBOL",            # Ensure gene identifiers are symbols
  ont = "BP",                    # Ontology (e.g., Biological Process)
  pAdjustMethod = "BH",
  pvalueCutoff = 0.07,
  qvalueCutoff = 0.2, 
  minGSSize = 5,
  maxGSSize = 500, 
  readable = TRUE
)

enrich_low_leftcolon <- enrichGO(
  gene = high_low_leftcolon$gene_symbol,   # DE gene list
  OrgDb = org.Hs.eg.db,          # Human gene annotation database
  keyType = "SYMBOL",            # Ensure gene identifiers are symbols
  ont = "BP",                    # Ontology (e.g., Biological Process)
  pAdjustMethod = "BH",
  pvalueCutoff = 0.07,
  qvalueCutoff = 0.2, 
  minGSSize = 5,
  maxGSSize = 500, 
  readable = TRUE
)
```
```{r}
dim(enrich_low_ileum)
dim(enrich_low_leftcolon)
head(enrich_low_ileum, 10)
head(enrich_low_leftcolon, 10)
```

## 2/08/2026 updated gseKEGG() instead of using enrichGO()
```{r}
geneList_ileum <- high_low_ileum %>%
  filter(!is.na(entrez), !is.na(stat)) %>%      # 只做 NA 过滤
  distinct(entrez, .keep_all = TRUE) %>%
  transmute(score = stat, entrez = as.character(entrez)) %>%
  { setNames(.$score, .$entrez) } %>%
  sort(decreasing = TRUE)

gsekegg_ileum <- gseKEGG(
  geneList = geneList_ileum,
  organism = "hsa",
  keyType  = "ncbi-geneid",
  eps = 0,
  ### added 08282026
  pvalueCutoff = 1,
  ###
  pAdjustMethod = "BH",
  minGSSize = 10,
  maxGSSize = 500,
  verbose = FALSE
)
dim(data.frame(gsekegg_ileum))
names(gsekegg_ileum@result)
```
```{r}
geneList_leftcolon <- high_low_leftcolon %>%
  filter(!is.na(entrez), !is.na(stat)) %>%      # 只做 NA 过滤
  distinct(entrez, .keep_all = TRUE) %>%
  transmute(score = stat, entrez = as.character(entrez)) %>%
  { setNames(.$score, .$entrez) } %>%
  sort(decreasing = TRUE)

gsekegg_leftcolon <- gseKEGG(
  geneList = geneList_leftcolon,
  organism = "hsa",
  keyType  = "ncbi-geneid",
  eps = 0,
  ### added 08282026
  pvalueCutoff = 1,
  ###
  pAdjustMethod = "BH",
  minGSSize = 10,
  maxGSSize = 500,
  verbose = FALSE
)
dim(data.frame(gsekegg_leftcolon))
```


### added 08282026 (to export the complete enrichment reuslts as supplementary)
```{r}
write.csv(
  as.data.frame(gsekegg_ileum),
  "/Users/guokexin/Documents/allenlee/manuscript/revision_072726/GSEA_ileum_SESCDle4_complete.csv",
  row.names = FALSE
)

write.csv(
  as.data.frame(gsekegg_leftcolon),
  "/Users/guokexin/Documents/allenlee/manuscript/revision_072726/GSEA_colon_SESCDle4_complete.csv",
  row.names = FALSE
)
```


## 6/25/2025 updated visualization
### heatmap
```{r}
library(ComplexHeatmap)
library(circlize)
library(enrichplot)
library(dplyr)
library(tidyr)
library(tibble)
library(grid)
# --- 1. Map terms to genes
term_gene_df <- enrich_low_ileum@result %>%
  dplyr::filter(p.adjust < 0.05) %>%
  dplyr::arrange(p.adjust) %>%
  dplyr::slice(1:15) %>%  # Top 15 terms
  dplyr::select(Description, geneID) %>%
  tidyr::separate_rows(geneID, sep = "/")
# --- 2. Merge with DE info
plot_df <- term_gene_df %>%
  left_join(high_low_ileum, by = c("geneID" = "gene_symbol")) %>%
  mutate(
    direction = ifelse(log2fc > 0, "UP", "DOWN"),
    log10padj = -log10(padj)
  )

# --- 3. Pivot to matrix form
mat_fc <- plot_df %>%
  dplyr::select(Description, geneID, log2fc) %>%
  pivot_wider(names_from = geneID, values_from = log2fc) %>%
  column_to_rownames("Description") %>%
  as.matrix()

mat_padj <- plot_df %>%
  dplyr::select(Description, geneID, log10padj) %>%
  pivot_wider(names_from = geneID, values_from = log10padj) %>%
  column_to_rownames("Description") %>%
  as.matrix()

head(mat_fc)
```
```{r}
# --- 4. Custom heatmap: color = direction, transparency = -log10(padj)
# Normalize padj for alpha scaling
padj_scaled <- mat_padj / max(mat_padj, na.rm = TRUE)
padj_scaled[is.na(padj_scaled)] <- 0

# Define color by sign of log2FC
col_fun <- colorRamp2(c(-3, 0, 3), c("blue", "white", "red"))

# Draw heatmap
ht = Heatmap(
  mat_fc,
  name = "log2FC",
  col = col_fun,
  na_col = "grey90",
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  # cell_fun = function(j, i, x, y, width, height, fill) {
  #   grid.rect(
  #     x = x, y = y, width = width, height = height,
  #     gp = gpar(
  #       fill = adjustcolor(fill, alpha.f = padj_scaled[i, j]),
  #       col = NA
  #     )
  #   )
  # },
  show_column_names = TRUE,
  show_row_names = TRUE,
  column_names_gp = gpar(fontsize = 10, rot = 45),
  row_names_gp = gpar(fontsize = 10),
  heatmap_legend_param = list(
    title = "Log2FC (adj. p-value < 0.05)",
    at = c(-3, 0, 3),
    labels = c("Down", "Neutral", "Up")
  )
)
library(grid)

draw(ht,
     heatmap_legend_side = "right",
     #annotation_legend_list = list(signif_legend),
     annotation_legend_side = "right",
     column_title = "High vs Low Pro – Ileum",
     column_title_gp = gpar(fontsize = 14, fontface = "bold"))

```
```{r}
## different threshold on p.adjusted for lc (higher)
# --- 1. Top enriched terms (left colon)
term_gene_df_lc <- enrich_low_leftcolon@result %>%
  dplyr::filter(p.adjust < 0.1) %>%
  dplyr::arrange(p.adjust)  %>%
  dplyr::slice(1:15) %>%  # Top 15 terms
  dplyr::select(Description, geneID) %>%
  tidyr::separate_rows(geneID, sep = "/")

# --- 2. Merge with DE info
plot_df_lc <- term_gene_df_lc %>%
  left_join(high_low_leftcolon, by = c("geneID" = "gene_symbol")) %>%
  mutate(
    direction = ifelse(log2fc > 0, "UP", "DOWN"),
    log10padj = -log10(padj)
  )

# --- 3. Pivot to matrix form
mat_fc_lc <- plot_df_lc %>%
  dplyr::select(Description, geneID, log2fc) %>%
  pivot_wider(names_from = geneID, values_from = log2fc) %>%
  column_to_rownames("Description") %>%
  as.matrix()

mat_padj_lc <- plot_df_lc %>%
  dplyr::select(Description, geneID, log10padj) %>%
  pivot_wider(names_from = geneID, values_from = log10padj) %>%
  column_to_rownames("Description") %>%
  as.matrix()

# --- 4. Normalize padj for transparency
padj_scaled_lc <- mat_padj_lc / max(mat_padj_lc, na.rm = TRUE)
padj_scaled_lc[is.na(padj_scaled_lc)] <- 0

# --- 5. Color scale
col_fun <- colorRamp2(c(-3, 0, 3), c("blue", "white", "red"))

# --- 6. Build heatmap
ht_lc <- Heatmap(
  mat_fc_lc,
  name = "log2FC",
  col = col_fun,
  na_col = "grey90",
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  # cell_fun = function(j, i, x, y, width, height, fill) {
  #   grid.rect(
  #     x = x, y = y, width = width, height = height,
  #     gp = gpar(
  #       fill = adjustcolor(fill, alpha.f = padj_scaled_lc[i, j]),
  #       col = NA
  #     )
  #   )
  # },
  show_column_names = TRUE,
  show_row_names = TRUE,
  column_names_gp = gpar(fontsize = 10, rot = 45),
  row_names_gp = gpar(fontsize = 10),
  heatmap_legend_param = list(
    title = "Log2FC (adj. p-value < 0.05)",
    at = c(-3, 0, 3),
    labels = c("Down", "Neutral", "Up")
  )
)

# --- 7. Draw with title
draw(ht_lc,
     heatmap_legend_side = "right",
     annotation_legend_side = "right",
     column_title = "High vs Low Pro – Left Colon",
     column_title_gp = gpar(fontsize = 14, fontface = "bold"))
```

### 2/08/2026 updated heatmap
```{r}
library(dplyr)
library(tidyr)
library(tibble)
library(ComplexHeatmap)
library(circlize)
library(grid)

plot_kegg_gsea_heatmap_symbol <- function(gsea_obj,
                                         de_tbl,
                                         site_title = "",
                                         top_terms = 15,
                                         padj_term_cutoff = 0.05,
                                         abs_log2fc_plot = NULL,   # e.g., 2; set NULL to skip
                                         min_gene_freq = 1,        # keep genes appearing in >= this many terms
                                         max_genes = 40) {

  # --- harmonize DE table columns
  if (!"log2fc" %in% colnames(de_tbl) && "log2FoldChange" %in% colnames(de_tbl)) {
    de_tbl <- de_tbl %>% rename(log2fc = log2FoldChange)
  }
  stopifnot(all(c("entrez", "log2fc", "padj") %in% colnames(de_tbl)))

  # --- 1) pathway -> entrez (core_enrichment)
  term_gene <- gsea_obj@result %>%
    filter(!is.na(p.adjust), p.adjust < padj_term_cutoff) %>%
    arrange(p.adjust) %>%
    slice_head(n = top_terms) %>%
    select(Description, core_enrichment) %>%
    separate_rows(core_enrichment, sep = "/") %>%
    rename(entrez = core_enrichment) %>%
    mutate(entrez = suppressWarnings(as.integer(entrez))) %>%
    filter(!is.na(entrez)) %>%
    distinct()

  if (nrow(term_gene) == 0) {
    stop("No pathways passed the p.adjust cutoff. Try increasing padj_term_cutoff or lowering top_terms.")
  }

  # --- 2) Entrez -> SYMBOL mapping
  map_df <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys = unique(as.character(term_gene$entrez)),
    keytype = "ENTREZID",
    columns = c("SYMBOL")
  ) %>%
    transmute(entrez = suppressWarnings(as.integer(ENTREZID)), symbol = SYMBOL) %>%
    filter(!is.na(entrez), !is.na(symbol)) %>%
    distinct()

  # --- 3) merge with DE info + SYMBOL
  plot_df <- term_gene %>%
    left_join(de_tbl, by = "entrez") %>%
    left_join(map_df, by = "entrez") %>%
    filter(!is.na(symbol), !is.na(log2fc), !is.na(padj)) %>%
    mutate(log10padj = -log10(padj))

  # OPTIONAL: abs(log2fc) filter for plotting only
  if (!is.null(abs_log2fc_plot)) {
    plot_df_f <- plot_df %>% filter(abs(log2fc) >= abs_log2fc_plot)
    if (nrow(plot_df_f) >= 20) {
      plot_df <- plot_df_f
    } else {
      message("Too few genes after abs(log2fc) filter; plotting without that filter.")
    }
  }

  # --- 4) handle duplicate SYMBOLs (multiple Entrez -> same symbol)
  # Strategy: keep the row with largest |log2fc| within each (Description, symbol)
  plot_df <- plot_df %>%
    group_by(Description, symbol) %>%
    slice_max(order_by = abs(log2fc), n = 1, with_ties = FALSE) %>%
    ungroup()

  # --- 5) build matrices (rows=pathways, cols=symbols)
  mat_fc <- plot_df %>%
    select(Description, symbol, log2fc) %>%
    pivot_wider(names_from = symbol, values_from = log2fc) %>%
    column_to_rownames("Description") %>%
    as.matrix()

  mat_padj <- plot_df %>%
    select(Description, symbol, log10padj) %>%
    pivot_wider(names_from = symbol, values_from = log10padj) %>%
    column_to_rownames("Description") %>%
    as.matrix()

  # --- 6) reduce columns for readability (optional)
  gene_freq <- colSums(!is.na(mat_fc))
  keep <- names(gene_freq[gene_freq >= min_gene_freq])
  if (length(keep) > 0) {
    mat_fc <- mat_fc[, keep, drop = FALSE]
    mat_padj <- mat_padj[, keep, drop = FALSE]
  }

  if (ncol(mat_fc) > max_genes) {
    gene_score <- apply(abs(mat_fc), 2, function(x) max(x, na.rm = TRUE))
    top_genes <- names(sort(gene_score, decreasing = TRUE))[1:max_genes]
    mat_fc <- mat_fc[, top_genes, drop = FALSE]
    mat_padj <- mat_padj[, top_genes, drop = FALSE]
  }

  # --- 7) heatmap
  col_fun <- colorRamp2(c(-3, 0, 3), c("blue", "white", "red"))

  ht <- Heatmap(
    mat_fc,
    name = "log2FC",
    col = col_fun,
    na_col = "grey90",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_column_names = TRUE,
    show_row_names = TRUE,
    column_names_gp = gpar(fontsize = 9, rot = 45),
    row_names_gp = gpar(fontsize = 10),
    heatmap_legend_param = list(
      title = "Log2FC (adj. p-value < 0.05)",
      at = c(-3, 0, 3),
      labels = c("Down", "Neutral", "Up")
    )
  )

  draw(
    ht,
    heatmap_legend_side = "right",
    annotation_legend_side = "right",
    column_title = site_title,
    column_title_gp = gpar(fontsize = 14, fontface = "bold")
  )

  invisible(list(mat_fc = mat_fc, mat_padj = mat_padj, plot_df = plot_df))
}
```
```{r}
# de table should be your FULL DESeq2 results for ileum
# must include: entrez, padj, log2FoldChange (or log2fc)
out_ileum <- plot_kegg_gsea_heatmap_symbol(
  gsea_obj = gsekegg_ileum,
  de_tbl = high_low_ileum,
  site_title = "qCD+S vs qCD-S – Ileum",
  top_terms = 10,
  padj_term_cutoff = 0.05,
  abs_log2fc_plot = 2,     # set NULL if you don't want this filter
  min_gene_freq = 1,
  max_genes = 15
)
out_leftcolon <- plot_kegg_gsea_heatmap_symbol(
  gsea_obj = gsekegg_leftcolon,
  de_tbl = high_low_leftcolon,
  site_title = "qCD+S vs qCD-S – Rectosigmoid Colon",
  top_terms = 10,
  padj_term_cutoff = 0.05,
  abs_log2fc_plot = 2,     # set NULL if you don't want this filter
  min_gene_freq = 1,
  max_genes = 15
)

```

### forest plot
```{r}
# Get top GO terms per site
top_go_ileum <- enrich_low_ileum@result %>%
  dplyr::arrange(p.adjust) %>%
  dplyr::slice(1:10)

top_go_colon <- enrich_low_leftcolon@result %>%
  dplyr::arrange(p.adjust) %>%
  dplyr::slice(1:10)

# Extract genes from these terms
get_genes_from_terms <- function(enrich_result) {
  enrich_result %>%
    dplyr::select(Description, geneID) %>%
    tidyr::separate_rows(geneID, sep = "/")
}

genes_ileum <- get_genes_from_terms(top_go_ileum)
genes_colon <- get_genes_from_terms(top_go_colon)

# Merge with DE results
genes_ileum_df <- genes_ileum %>%
  left_join(high_low_ileum, by = c("geneID" = "gene_symbol")) %>%
  mutate(site = "Ileum")

genes_colon_df <- genes_colon %>%
  left_join(high_low_leftcolon, by = c("geneID" = "gene_symbol")) %>%
  mutate(site = "Left Colon")

# Combine
forest_df <- bind_rows(genes_ileum_df, genes_colon_df) %>%
  dplyr::filter(!is.na(log2fc)) %>%
  mutate(
    gene_term = paste(geneID, "\n(", Description, ")"),
    lfcSE = ifelse(is.na(lfcSE), 0.2, lfcSE),  # use actual SE if available
    lower = log2fc - 1.96 * lfcSE,
    upper = log2fc + 1.96 * lfcSE
  )

```
```{r}
library(ggplot2)
# Choose N
top_n <- 15

# Select top N genes per site by abs(log2FC)
forest_df_limited <- forest_df %>%
  group_by(site) %>%
  arrange(padj) %>%
  slice_head(n = top_n) %>%
  ungroup()
ggplot(forest_df_limited, aes(x = log2fc, y = reorder(gene_term, log2fc), color = ifelse(log2fc > 0, "Upregulated", "Downregulated"))) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  facet_wrap(~site, scales = "free_y") +
  scale_color_manual(values = c("Downregulated" = "#1f77b4", "Upregulated" = "#d62728"), name = "Direction") +
  labs(
    x = "log2 Fold Change (95% CI)",
    y = "Gene (GO Term)",
    title = "Forest Plot of top DE Genes in Enriched GO Terms"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    strip.text = element_text(face = "bold")
  )
```
```{r}
ggplot(forest_df, aes(
  x = log2fc,
  y = reorder(geneID, log2fc),
  color = site
)) +
  geom_point(size = 2, position = position_dodge(width = 0.5)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 height = 0.2,
                 position = position_dodge(width = 0.5)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("Ileum" = "#1f77b4", "Left Colon" = "#d62728")) +
  labs(
    x = "log2 Fold Change (95% CI)",
    y = "Gene",
    title = "Comparison of DE Genes Between Sites",
    color = "Site"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    axis.text.y = element_text(size = 10)
  )

ggplot(forest_df_limited, aes(x = log2fc, y = reorder(gene_term, log2fc), color = site)) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  #facet_wrap(~site, scales = "free_y") +
  scale_color_manual(values = c("Ileum" = "#1f77b4", "Left Colon" = "#d62728")) +
  labs(
    x = "log2 Fold Change (95% CI)",
    y = "Gene (GO Term)",
    title = "Forest Plot of top DE Genes in Enriched GO Terms"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    strip.text = element_text(face = "bold")
  )
```
### 2/08/2026 updated forest plot using kegg()
```{r}
## ---- A) choose top pathways per site (by p.adjust) ----
top_kegg_ileum <- gsekegg_ileum@result %>%
  arrange(p.adjust) %>%
  slice_head(n = 10)

top_kegg_colon <- gsekegg_leftcolon@result %>%
  arrange(p.adjust) %>%
  slice_head(n = 10)

## ---- B) extract leading-edge genes from pathways ----
get_entrez_from_gsea_terms <- function(gsea_term_df) {
  gsea_term_df %>%
    select(Description, core_enrichment) %>%
    separate_rows(core_enrichment, sep = "/") %>%
    rename(entrez = core_enrichment) %>%
    mutate(entrez = suppressWarnings(as.integer(entrez))) %>%
    filter(!is.na(entrez)) %>%
    distinct()
}

genes_ileum <- get_entrez_from_gsea_terms(top_kegg_ileum) %>% mutate(site = "Ileum")
genes_colon <- get_entrez_from_gsea_terms(top_kegg_colon) %>% mutate(site = "Rectosigmoid Colon")

## ---- C) map Entrez -> SYMBOL for labels ----
all_entrez <- unique(c(genes_ileum$entrez, genes_colon$entrez))

entrez2sym <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = as.character(all_entrez),
  keytype = "ENTREZID",
  columns = c("SYMBOL")
) %>%
  transmute(entrez = suppressWarnings(as.integer(ENTREZID)), gene_symbol = SYMBOL) %>%
  filter(!is.na(entrez), !is.na(gene_symbol)) %>%
  distinct()

## ---- D) choose DE tables (prefer FULL DESeq2 results tables) ----
## These should have: entrez, padj, log2fc (or log2FoldChange), lfcSE (optional but good)
de_ileum <- high_low_ileum
de_colon <- high_low_leftcolon

if (!"log2fc" %in% colnames(de_ileum) && "log2FoldChange" %in% colnames(de_ileum)) {
  de_ileum <- de_ileum %>% rename(log2fc = log2FoldChange)
}
if (!"log2fc" %in% colnames(de_colon) && "log2FoldChange" %in% colnames(de_colon)) {
  de_colon <- de_colon %>% rename(log2fc = log2FoldChange)
}

## ---- E) merge genes with DE results ----
genes_ileum_df <- genes_ileum %>%
  left_join(entrez2sym, by = "entrez") %>%
  left_join(de_ileum, by = "entrez")

genes_colon_df <- genes_colon %>%
  left_join(entrez2sym, by = "entrez") %>%
  left_join(de_colon, by = "entrez")

forest_df <- bind_rows(genes_ileum_df, genes_colon_df) %>%
  filter(!is.na(log2fc), !is.na(gene_symbol.x)) %>%
  mutate(
    gene_term = paste0(gene_symbol.x, "\n(", Description, ")"),
    lfcSE = ifelse(is.na(lfcSE), 0.2, lfcSE),  # fallback only if needed
    lower = log2fc - 1.96 * lfcSE,
    upper = log2fc + 1.96 * lfcSE
  )
```
```{r}
library(ggplot2)
top_n <- 5

forest_df_limited <- forest_df %>%
  mutate(gene_symbol = gene_symbol.x) %>%                 # your current column name
  filter(!is.na(gene_symbol), !is.na(padj), !is.na(log2fc)) %>%
  group_by(site, gene_symbol) %>%
  arrange(padj, desc(abs(log2fc)), .by_group = TRUE) %>%  # choose representative pathway per gene
  slice(1) %>%
  ungroup() %>%
  group_by(site) %>%
  arrange(padj, desc(abs(log2fc)), .by_group = TRUE) %>%  # then pick top genes per site
  slice_head(n = top_n) %>%
  ungroup() %>%
  mutate(
    gene_term = paste0(gene_symbol, "\n(", Description, ")")  # re-make label after dedup
  )

ggplot(
  forest_df_limited,
  aes(
    x = log2fc,
    y = reorder(gene_term, log2fc),
    color = ifelse(log2fc > 0, "Upregulated", "Downregulated")
  )
) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  facet_wrap(~site, scales = "free_y") +
  scale_color_manual(
    values = c("Downregulated" = "#1f77b4", "Upregulated" = "#d62728"),
    name = "Direction"
  ) +
  labs(
    x = "log2 Fold Change (95% CI)",
    y = "Gene (KEGG Pathway)",
    title = "Forest Plot of Leading-edge DE Genes in Top KEGG Pathways"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    strip.text = element_text(face = "bold")
  )

```



## original visualization
```{r}
### high vs low in ileum
## Dotplot 
dotplot(enrich_low_ileum, showCategory=15, size = "Count")
#barplot(enrich_results_go, showCategory = 20) 
# concept
cnetplot(enrich_low_ileum, showCategory = 10)
# Heatplot
heatplot(enrich_low_ileum)
## Enrichmap clusters the xx most significant (by padj) GO terms to visualize relationships between terms
enrich_low_ileum_pair = pairwise_termsim(enrich_low_ileum)
emapplot(enrich_low_ileum_pair, showCategory = 10)
```
```{r}
### high vs low in left colon
## Dotplot 
dotplot(enrich_low_leftcolon, showCategory=15, size = "Count")
#barplot(enrich_results_go, showCategory = 20) 
# concept
cnetplot(enrich_low_leftcolon, showCategory = 10)
# Heatplot
heatplot(enrich_low_leftcolon)
## Enrichmap clusters the xx most significant (by padj) GO terms to visualize relationships between terms
enrich_low_leftcolon_pair = pairwise_termsim(enrich_low_leftcolon)
emapplot(enrich_low_leftcolon_pair, showCategory = 10)
```

```{r}
enrich_active_ileum <- enrichGO(
  gene = high_active_ileum$gene_symbol,   # DE gene list
  OrgDb = org.Hs.eg.db,          # Human gene annotation database
  keyType = "SYMBOL",            # Ensure gene identifiers are symbols
  ont = "BP",                    # Ontology (e.g., Biological Process)
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2, 
  minGSSize = 10,
  maxGSSize = 500, 
  readable = TRUE
)

enrich_active_leftcolon <- enrichGO(
  gene = high_active_leftcolon$gene_symbol,   # DE gene list
  OrgDb = org.Hs.eg.db,          # Human gene annotation database
  keyType = "SYMBOL",            # Ensure gene identifiers are symbols
  ont = "BP",                    # Ontology (e.g., Biological Process)
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2, 
  minGSSize = 10,
  maxGSSize = 500, 
  readable = TRUE
)
```
```{r}
### high vs active in ileum
## Dotplot 
dotplot(enrich_active_ileum, showCategory=15, size = "Count")
#barplot(enrich_results_go, showCategory = 20) 
# concept
cnetplot(enrich_active_ileum, showCategory = 10)
# Heatplot
heatplot(enrich_active_ileum)
## Enrichmap clusters the xx most significant (by padj) GO terms to visualize relationships between terms
enrich_active_ileum_pair = pairwise_termsim(enrich_active_ileum)
emapplot(enrich_active_ileum_pair, showCategory = 10)
```

```{r}
### high vs active in left colon
## Dotplot 
dotplot(enrich_active_leftcolon, showCategory=15, size = "Count")
#barplot(enrich_results_go, showCategory = 20) 
# concept
cnetplot(enrich_active_leftcolon, showCategory = 10)
# Heatplot
heatplot(enrich_active_leftcolon)
## Enrichmap clusters the xx most significant (by padj) GO terms to visualize relationships between terms
enrich_active_leftcolon_pair = pairwise_termsim(enrich_active_leftcolon)
emapplot(enrich_active_leftcolon_pair, showCategory = 10)
```


```{r}
### left colon are all unmatched genes
# library(AnnotationDbi)  # Provides mapping functions
# library(EnsDb.Hsapiens.v86)
# mapped_genes <- mapIds(EnsDb.Hsapiens.v86,
#                        keys = high_low_leftcolon$gene_symbol,
#                        column = "ENTREZID",
#                        keytype = "SYMBOL",
#                        multiVals = "first")  # 'first' chooses the first mapping if multiple exist
# mapped_genes
```
```{r}
# # Remove NA (unmapped) genes
# mapped_genes <- na.omit(mapped_genes)
# 
# enrich_results_leftcolon <- enrichGO(
#   gene = mapped_genes,   # DE gene list
#   OrgDb = org.Hs.eg.db,          # Human gene annotation database
#   keyType = "ENTREZID",            # Ensure gene identifiers are symbols
#   ont = "BP",                    # Ontology (e.g., Biological Process)
#   pAdjustMethod = "BH",
#   pvalueCutoff = 0.05,
#   qvalueCutoff = 0.2,
#   minGSSize = 10,
#   maxGSSize = 500,
#   readable = TRUE
# )
```
```{r}
# library(DECIPHER)
```

```{r}
### prep for gseGO (ranked by log2foldchange)
# gene_list_leftcolon <- high_low_leftcolon$log2fc
# names(gene_list_leftcolon) <- high_low_leftcolon$gene_symbol
# 
# # Sort the genes by log2 fold change for ranking (required for GSEA)
# gene_list_leftcolon <- sort(gene_list_leftcolon, decreasing = TRUE)
# gene_list_leftcolon
```
```{r}
# gsea_results_ileum <- gseGO(
#   geneList = gene_list_leftcolon,         # Ranked gene list
#   OrgDb = org.Hs.eg.db,               # Gene annotation database
#   keyType = "SYMBOL",                 # Gene identifier (symbols in your case)
#   ont = "BP",                         
#   minGSSize = 10,                     # Minimum gene set size
#   maxGSSize = 500,                    # Maximum gene set size
#   pvalueCutoff = 0.05,                # P-value threshold
#   pAdjustMethod = "BH",               # Adjust p-values with Benjamini-Hochberg
#   verbose = TRUE
# )

```


```{r}
# ## Dotplot 
# dotplot(enrich_results_leftcolon, showCategory=15, size = "Count")
# #barplot(enrich_results_go, showCategory = 20) 
# # concept
# cnetplot(enrich_results_leftcolon, showCategory = 10)
# # Heatplot
# heatplot(enrich_results_leftcolon)
# ## Enrichmap clusters the xx most significant (by padj) GO terms to visualize relationships between terms
# enrich_results_leftcolon_pair = pairwise_termsim(enrich_results_leftcolon)
# emapplot(enrich_results_leftcolon_pair, showCategory = 10)
```
